---
title: 'Class 6: Partial pooling, zero-inflation, and multinomial models'
author: Andrew Parnell \newline \texttt{andrew.parnell@ucd.ie} \newline \vspace{1cm}
  \newline \includegraphics[width=1cm]{UCDlogo.pdf}
output:
  beamer_presentation:
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')
#knitr::opts_knit$set(global.par = TRUE)
setwd("~/GitHub/bhm_course/slides")
set.seed(123)
```
```{r, include=FALSE}
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Learning outcomes:

- Be able to describe the advantages of partial pooling
- Be able to fit some basic zero inflation and hurdle models
- Be able to understand and fit some multinomial modelling examples

\vspace{2cm}

- For this lecture we are going to switch to Stan

## A false dichotomy: fixed vs random effects

- We've been fitting a model with varying intercepts and slopes to the earnings data:
$$y_i \sim N(\alpha_{\mbox{eth}_i} + \beta_{\mbox{eth}_i} x_i, \sigma^2)$$
where:
$$\alpha_j \sim N(\mu_\alpha, \sigma_\alpha^2) \mbox{ and } \beta_j \sim N(\mu_\beta, \sigma_\beta^2)$$
- In traditional parlance this is a random effects model
- When we fit our model we are learning about the values of the slopes and intercepts, and also the values of their means and standard deviations

## The extremes of varying vs fixed parameters

- Now consider what happens when $\sigma_\alpha$ and $\sigma_\beta$ get smaller and smaller. What will happen to the values of the slopes and the intercepts?
- Alternatively, consider what happens as $\sigma_\alpha$ and $\sigma_\beta$ get larger and larger?
- Are these still random effects models?

## The advantages of borrowing strength

- The process of $\sigma_\alpha$ and $\sigma_\beta$ getting smaller or larger will control the degree to which the slopes and intercepts are similar to each other
- If they are similar to each other we say they are _borrowing strength_ as data in the other groups is influencing the intercept/slope. This is a powerful idea
- Mathematically you can write out the estimated mean of the parameters as a weighted average of the group mean and the overall mean where the weights are dependent on the group and overall variance and sample sizes. 
- Because of the weighted nature of the estimate this is often called _partial pooling_

## Zero-inflation and hurdle models

- Let's introduce some new data. This is data from an experiment on whiteflies:

```{r}
wf = read.csv('../data/whitefly.csv')
head(wf)
```

The response variable here is the count `imm` of immature whiteflies, and the explanatory variables are `block` (plant number), `week`, and  treatment `treat`.

## Look at those zeros!

```{r}
barplot(table(wf$imm), 
        main = 'Number of immature whiteflies')
```

## A first model

- These are count data so a Poisson distribution is a good start
- Let's consider a basic Poisson distribution model for $Y_{i}$, $i= 1, \ldots, N$ observations:
$$Y_i \sim Po(\lambda_i)$$
$$\log(\lambda_i) = \beta_{\mbox{trt}_i}$$
- We'll only consider the treatment effect but we could run much more complicated models with e.g. other covariates and interactions

## Fitting the model in Stan

\tiny
```{r}
stan_code = '
data {
  int<lower=0> N;
  int<lower=0> N_trt;
  int<lower=0> y[N];
  int trt[N];
}
parameters {
  real beta_trt[N_trt];
  real trt_mean;
  real<lower=0> trt_sd;
}
model {
  for (i in 1:N) 
    y[i] ~ poisson_log(beta_trt[trt[i]]);

  # Priors on coefficients
  for(j in 1:N_trt)
    beta_trt[j] ~ normal(trt_mean, trt_sd);

  trt_mean ~ normal(0, 10);
  trt_sd ~ cauchy(0, 5);
}
'
```

## Running the model

```{r, message=FALSE, results='hide'}
stan_run = stan(data = list(N = nrow(wf), 
                            N_trt = length(unique(wf$trt)),
                            y = wf$imm,
                            trt = wf$trt),
                model_code = stan_code)
```

## Results

```{r, fig.height=5}
plot(stan_run)
```

Some clear treatment effects - treatment 5 in particular

## Did the model actually fit well?

```{r, fig.height=4}
pars = extract(stan_run, pars = 'beta_trt')$beta_trt
beta_means = apply(pars,2,'mean')
y_sim_mean = exp(beta_means[wf$trt])
y_sim = rpois(nrow(wf), y_sim_mean)
hist(wf$imm, breaks = seq(0,max(wf$imm)))
hist(y_sim, breaks = seq(0,max(wf$imm)), 
     add = TRUE, col = 'gray')
```

## What about the zeros?

- One way of broadening the distribution is through over-dispersion which we have already met:
$$\log(\lambda_i) \sim N(\beta_{\mbox{trt}_i}, \sigma^2) $$

- However this doesn't really solve the problem of excess zeros

- Instead there are a specific class of models called _zero-inflation_ models which use a specific probability distribution. The zero-inflated Poisson (ZIP) with ZI parameter $q_0$ is written as:

$$p(y|\lambda) = \left\{ \begin{array}{ll} q_0 + (1-q_0) \times Poisson(0, \lambda) & \mbox{if } y=0 \\ (1-q_0) \times Poisson(y, \lambda) & \mbox{if } y=0 \end{array} \right.$$

## Fitting models with custom probability distributions

- The Zero-inflated Poisson distribution is not included in Stan or JAGS by default. We have to create it
- It's pretty easy to create new probability distributions in Stan
- It's quite hard to do so in JAGS, requiring some weird tricks
- In Stan all of you have to do is give it a way of computing the likelihood score, which it keeps track of via a variable called `target`

## Fitting the ZIP in Stan

\tiny
```{r}
stan_code = '
data {
  int<lower=0> N;
  int<lower=0> N_trt;
  int<lower=0> y[N];
  int trt[N];    
}
parameters {
  real<lower=0, upper=1> q_0;
  real beta_trt[N_trt];
  real trt_mean;
  real<lower=0> trt_sd;
}
model {
  for(j in 1:N_trt)
    beta_trt[j] ~ normal(trt_mean, trt_sd);
  trt_mean ~ normal(0, 10);
  trt_sd ~ cauchy(0, 5);

  for (i in 1:N) {
    if (y[i] == 0)
      target += log_sum_exp(bernoulli_lpmf(1 | q_0),
                                bernoulli_lpmf(0 | q_0)
                                  + poisson_log_lpmf(y[i] | beta_trt[trt[i]]));
    else
      target += bernoulli_lpmf(0 | q_0) + poisson_log_lpmf(y[i] | beta_trt[trt[i]]);
  }
}
'
```

## Running the model

```{r, message=FALSE, results='hide'}
stan_run = stan(data = list(N = nrow(wf), 
                            N_trt = length(unique(wf$trt)),
                            y = wf$imm,
                            trt = wf$trt),
                model_code = stan_code)
```

## Results

```{r, fig.height=5}
plot(stan_run)
```

## Did it work any better? - code

```{r}
beta_means = apply(extract(stan_run, pars = 'beta_trt')$beta_trt,2,'mean')
q_0_mean = mean(extract(stan_run, pars = 'q_0')$q_0)
y_sim_mean = exp(beta_means[wf$trt])
rZIP = function(mean, q_0) {
  pois = rpois(length(mean), mean)
  pois[runif(length(mean))<q_0] = 0
  return(pois)
}
y_sim = rZIP(y_sim_mean, q_0_mean)
```

## Did it work any better? - picture
```{r, fig.height=5}
hist(wf$imm, breaks = seq(0,max(wf$imm)))
hist(y_sim, breaks = seq(0,max(wf$imm)), 
     add = TRUE, col = rgb(0.75,0.75,0.75,0.4))
```

## Some more notes on Zero-inflated Poisson

- This model seems to over-predict the number of zeros! It would be interesting to perhaps try having a different probability of zeros ($q_0$) for different treatments
- It might be that the other covariates explain some of the zero behaviour
- We could further add in both zero-inflation and over-dispersion

## An alternative: hurdle models

- ZI models work by having a parameter (here $q_0$) which is the probability of getting a zero, and so the probability of getting a Poisson value (which could also be a zero) is 1 minus this value
- An alternative (which is slightly more complicated) is a hurdle model where $q_0$ represents the probability of the _only way_ of getting a zero. With probability (1-$q_0$) we end up with a special Poisson random variable which has to take values 1 or more
- In some ways this is richer than a ZI model since zeros can be deflated or inflated

## A hurdle-Poisson model in Stan

\tiny
```{r}
stan_code = '
data {
  int<lower=0> N;
  int<lower=0> N_trt;
  int<lower=0> y[N];
  int trt[N];    
}
parameters {
  real<lower=0, upper=1> q_0;
  real beta_trt[N_trt];
  real trt_mean;
  real<lower=0> trt_sd;
}
model {
  for(j in 1:N_trt)
    beta_trt[j] ~ normal(trt_mean, trt_sd);
  trt_mean ~ normal(0, 10);
  trt_sd ~ cauchy(0, 5);

  for (i in 1:N) {
    if (y[i] == 0)
          target += log(q_0);
        else
          target += log1m(q_0) + poisson_log_lpmf(y[i] | beta_trt[trt[i]])
                    - poisson_lccdf(0 | exp(beta_trt[trt[i]]));
  }
}
'
```

## Running the model

```{r, message=FALSE, results='hide'}
stan_run = stan(data = list(N = nrow(wf), 
                            N_trt = length(unique(wf$trt)),
                            y = wf$imm,
                            trt = wf$trt),
                model_code = stan_code)
```

## Results

\tiny
```{r, fig.height = 5}
plot(stan_run)
```

## Did it work any better? - code

```{r}
beta_means = apply(extract(stan_run, pars = 'beta_trt')$beta_trt,2,'mean')
q_0_mean = mean(extract(stan_run, pars = 'q_0')$q_0)
y_sim_mean = exp(beta_means[wf$trt])
rZIP = function(mean, q_0) {
  pois = rpois(length(mean), mean)
  pois[runif(length(mean))<q_0] = 0
  return(pois)
}
y_sim = rZIP(y_sim_mean, q_0_mean)
```

## Did it work any better? - picture

```{r, fig.height=5}
hist(wf$imm, breaks = seq(0,max(wf$imm)))
hist(y_sim, breaks = seq(0,max(wf$imm)), 
     add = TRUE, col = rgb(0.75,0.75,0.75,0.4))
```


## The multinomial distribution

- Multinomial data can be thought of as multivariate discrete data
- It's usually used in two different scenarios:

    1. For classification, when you have an observation falling into a single one of $K$ possible categories
    1. For multinomial regression, where you have a set of counts which sum to a known value $N$
    
- We will just consider the multinomial regression case, whereby we have observations $y_i = \left[ y_{i1}, \ldots, y_{iK} \right]$ where the sum $\sum_{k=1}^K y_{ik} = N_i$ is fixed

- The classification version is a simplification of the regression version

## Some new data! - pollen

```{r}
pollen = read.csv('../data/pollen.csv')
head(pollen)
```

These data are pollen counts of 7 varieties of pollen from modern samples with two covariates

## Some plots

- The two covariates represent the length of the growing season (GDD5) and harshness of the winter (MTCO)
- The task is to find which climate regimes each pollen variety favours
```{r, fig.height = 5, echo = FALSE}
N = rowSums(pollen[,3:ncol(pollen)])
par(mfrow=c(1,2))
plot(pollen$GDD5, pollen$Pinus.D)
plot(pollen$GDD5, pollen$Pinus.D/N)
par(mfrow=c(1,1))
```

## A multinomial model

- The multinomial distribution is often written as:
$$\left[ y_{i1}, \ldots, y_{iK} \right] \sim Mult(N_i, \left\{ p_{i1}, \ldots, p_{iK} \right\})$$
or, for short:
$$y_i \sim Mult(N_i, p_i)$$

- The key parameters here are the probability vectors $p_i$. It's these we want to use a link function on to include the covariates

- We need to be careful as each must sum to one: $\sum_{k=1}^K p_{ik} = 1$. Any link function must satisfy this constraint

## Prior distributions on probability vectors

- When $K=2$ we're back the binomial-logit we met in the first day, and we can use the logit link function
- When $K>2$ a common function to use is the _soft-max_ function:
$$p_{ik} = \frac{ \exp(\theta_{ik})}{ \sum_{j=1}^K \exp(\theta_{ij})}$$
- This is a generalisation of the logit function
- The next layer of our model sets, e.g.:
$$\theta_{ik} = \alpha_k + \beta_k \mbox{GDD5} + \gamma_k \mbox{MTCO}$$

## Stan code part 1

\tiny
```{r}
stan_code = '
data {
  int<lower=1> n;
  int<lower=1> K;
  int<lower=0> y[n,K];
  real x1[n];
  real x2[n];
}
parameters {
  vector[K] alpha; 
  vector[K] beta;
  vector[K] gamma;
  real alpha_mean;
  real beta_mean;
  real gamma_mean;
  real<lower=0> sigma_alpha;
  real<lower=0> sigma_beta;
  real<lower=0> sigma_gamma;
}
transformed parameters {
  vector[K] theta[n];
  simplex[K] p[n]; 
  
  for(i in 1:n){
    theta[i] =  alpha + beta*x1[i] + gamma*x2[i];
  }
  for(i in 1:n){
    p[i] = softmax(theta[i]);
  }
}
...
'
```

## Stan code part 2

\tiny
```{r}
stan_code = '
...
model {
  for(k in 1:K) {
    alpha[k] ~ normal(alpha_mean,sigma_alpha);
    beta[k] ~ normal(beta_mean,sigma_beta);
    gamma[k] ~ normal(gamma_mean,sigma_gamma);
  }
  alpha_mean ~ normal(0, 10);
  beta_mean ~ normal(0, 10);
  gamma_mean ~ normal(0, 10);
  sigma_alpha ~ cauchy(0, 5);
  sigma_beta ~ cauchy(0, 5);
  sigma_gamma ~ cauchy(0, 5);

  for(i in 1:n)
    y[i] ~ multinomial(p[i]);
}
'
```


```{r, eval=FALSE}
# pollen2 = pollen[sample(1:nrow(pollen), 200),]
# stan_run = stan(data = list(K = 3,
#                             n = nrow(pollen2), 
#                             y = pollen2[,3:5],
#                             x1 = scale(pollen2$GDD5)[,1],
#                             x2  = scale(pollen2$MTCO)[,1]),
#                 model_code = stan_code)
```

## Notes about this model

- This model is not going to fit very well, since it is unlikely that a linear relationship between the covariates and the pollen counts will match the data
- It might be better to use e.g. a spline model (not covered in this course, but we can talk about it)
- Similarly we might need some complex interactions between the covariates as they are strongly linked
- We have constrained the parameters here so that the slopes and intercepts borrow strength across species. Does this make sense? What else could we do?

## Some final notes about multinomial models

- These models can be a pain to deal with as there are tricky constraints on the $\theta$ parameters to make them all sum to 1. 
- The `softmax` function is one choice but there are lots of others (logistic ratios, the Dirichlet distribution, ...)
- Whilst the classification version of this model just has binary $y_i$ (with just a single 1 in it) most packages (including JAGS and Stan) have a special distribution (e.g. `dcat` in JAGS) for this situation

## Summary

- We have seen how partial pooling is a balance between a model of complete independence and complete dependence between groups
- We have fitted some zero inflated and hurdle Poisson models
- We have seen some (poorly fitting) multinomial regression models
