---
title: 'Class 6: Hierarchical generalised linear models'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{maynooth_uni_logo.jpg}
output:
  beamer_presentation:
    includes:
      in_header: header.tex
editor_options: 
  chunk_output_type: console
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
```

## Learning outcomes:

- Understand the modelling implications of moving from linear to hierarchical generalised linear models (HGLMs)
- Know some of the different versions of Hierarchical GLMs
- Be able to fit HGLMS in JAGS
- Be able to expand and summarise fitted models

## From LMs to HGLMs

- Reminder: a hierarchical model has prior distributions on the parameters which depend on further parameters
- A generalised linear model is one in which the probability distribution is not normal, and a link function serves to match the mean of the distribution to the covariates
- Within this framework, we can borrow the ideas from the previous class to create hierarchical GLMs
- We will go through four examples: binomial-logit, Poisson, robust regression, and ordinal regression 

## Example 1: binomial-logit

- In class 2, we met the Binomial-logit model for binary data:
$$y_i \sim Bin(1, p_i), logit(p_i) = \alpha + \beta (x_i - \bar{x})$$
Here $logit(p_i)$ is the link function equal to $\log \left( \frac{p_i}{1-p_i} \right)$ and transforms the bounded probabilities into an unbounded space

- If we have non-binary data we just change the likelihood:
$$y_i \sim Bin(N_i, p_i), logit(p_i) = \alpha + \beta (x_i - \bar{x})$$

- In a hierarchical version of this model, we vary the _latent parameters_ $\alpha$ and $\beta$ and give them prior distributions

## The swiss willow tit data

\tiny
```{r}
swt = read.csv('../data/swt.csv')
head(swt)
```

## A hierarchical model

\small
- Suppose we want to fit a model on the sum $y_i =$ `rep.1 + rep.2 + rep.3`:
$$y_i \sim Bin(N_i, p_i), logit(p_i) = \alpha_{\mbox{altitude}_i} + \beta_{\mbox{altitude}_i} (x_i- \bar{x})$$
where $x_i$ is the percentage of forest cover

- What prior distributions should we use for $\alpha$ and $\beta$?

- Useful side note: A value of 10 on the logit scale leads to a probability of about 1, and a value of -10 leads to a probability of about 0 (you can test this by typing `inv.logit(10)`) so I wouldn't expect the value of $logit(p_i)$ to ever get much bigger than 10 or smaller than -10

- I have no idea whether we are more likely to find these birds in high percentage forest or low, so I'm happy to think that $\beta$ might be around zero, and be positive or negative. Forest cover ranges from 0 to 100 so that suggests that $\beta$ is very unlikely to be bigger than 0.1 or smaller than -0.1. Perhaps $\beta \sim N(0, 0.1^2)$ is a good prior

- It looks to me like the intercept is very unlikely to be outside the range (-10, 10) so perhaps $\alpha \sim N(0, 5^2)$ is appropriate

## JAGS code

\small
```{r, message = FALSE, results = 'hide'}
jags_code = '
model{
  # Likelihood
  for(i in 1:N) {
    y[i] ~ dbin(p[i], N_exp[i])
    logit(p[i]) <- alpha[alt[i]] + beta[alt[i]]* (x[i] - mean(x))
  }
  # Priors
  for(j in 1:N_alt) {
    alpha[j] ~ dnorm(mu_alpha, sigma_alpha^-2)
    beta[j] ~ dnorm(mu_beta, sigma_beta^-2)
  }
  mu_alpha ~ dnorm(0, 5^-2)
  mu_beta ~ dnorm(0, 0.1^-2)
  sigma_alpha ~ dt(0,5,1)T(0,)
  sigma_beta ~ dt(0,5,1)T(0,)
}
'
```

## Model fit - intercepts

```{r, echo = FALSE, message=FALSE, results = 'hide'}
sum_fun = function(x) {
  s = ifelse(is.na(x[1]),0,x[1]) + ifelse(is.na(x[2]),0,x[2]) + ifelse(is.na(x[3]),0,x[3])
  N = ifelse(is.na(x[1]),0,1) + ifelse(is.na(x[2]),0,1) + ifelse(is.na(x[3]),0,1)
  return(c(s,N))
}
y = apply(swt[,1:3],1,sum_fun)[1,]
N = apply(swt[,1:3],1,sum_fun)[2,]
library(R2jags)
jags_run = jags(data = list(N = nrow(swt),
                            N_exp = N,
                            N_alt = length(unique(swt$alt)),
                            alt = swt$alt,
                            y = y,
                            x = swt$forest),
                parameters.to.save = c('alpha',
                                       'beta',
                                       'mu_alpha',
                                       'mu_beta',
                                       'sigma_alpha',
                                       'sigma_beta'),
                model.file = textConnection(jags_code))
```
```{r, echo = FALSE, fig.height=5}
par(mfrow=c(1,3))
pars = jags_run$BUGSoutput$sims.list
for(i in c(2,3,1)) {
  hist(pars$alpha[,i], breaks = 30, main = paste('Altitude type:',levels(swt$alt)[i]), xlim = range(pars$alpha), xlab = 'Intercept value')  
}
par(mfrow=c(1,1))
```

## Model fit - Slopes

```{r, echo = FALSE, fig.height=5}
par(mfrow=c(1,3))
pars = jags_run$BUGSoutput$sims.list
for(i in c(2,3,1)) {
  hist(pars$beta[,i], breaks = 30, 
       main = paste('Altitude type:',levels(swt$alt)[i]), 
       xlim = range(pars$beta), xlab = 'Slope value')  
}
par(mfrow=c(1,1))
```

## Model fit - estimated mean proportions

```{r, echo=FALSE, fig.height=5}
library(boot)
par(mfrow=c(1,3))
pars = jags_run$BUGSoutput$mean
for(i in c(2,3,1)) {
  plot(swt$forest, y/N, main = paste('Altitude type:',levels(swt$alt)[i]), xlab = '% forest cover', ylab = 'Estimated proporton')
  points(swt$forest, inv.logit(pars$alpha[i] + pars$beta[i] * swt$forest), col = i+1)
}
par(mfrow=c(1,1))
```

## Type 2: Poisson HGLMs

- For a Poisson distribution there is no upper bound on the number of counts

- We just change the likelihood (to Poisson) and the link function (to $\log$):
$$y_i \sim Po(\lambda_i), \log(\lambda_i) = \alpha + \beta (x_i - \bar{x}))$$

- We can now add our hierarchical layers into $\alpha$ and $\beta$, or...

- Another way we can add an extra layer is by giving $\log(\lambda_i)$ a probability distribution rather than setting it to a value

- This is a way of introducing _over-dispersion_, i.e. saying that the data are more variable than that expected by a standard Poisson distribution with our existing covariates

## An over-dispersed model

- The over-dispersed model looks like:
$$y_i \sim Po(\lambda_i), \log(\lambda_i) \sim N(\alpha + \beta (x_i - \bar{x}), \sigma^2)$$
where $\sigma$ is the over-dispersion parameter

- We now need to estimate prior distributions for $\alpha$, $\beta$, and $\sigma$

- We will use the SWT data again, but pretend that we didn't know that they had gone out $N$ times looking for the birds

## JAGS code for OD Poisson

```{r, message = FALSE, results = 'hide'}
jags_code = '
model{
  # Likelihood
  for(i in 1:N) {
    y[i] ~ dpois(exp(log_lambda[i]))
    log_lambda[i] ~ dnorm(alpha + beta * (x[i] - mean(x)), 
        sigma^-2)
  }
  alpha ~ dnorm(0, 5^-2)
  beta ~ dnorm(0, 0.1^-2)
  sigma ~ dt(0,5,1)T(0,)
}
'
```

## Model run

```{r, echo = FALSE, message=FALSE, results = 'hide'}
set.seed(123)
jags_run = jags(data = list(N = nrow(swt),
                            y = y,
                            x = swt$forest),
                parameters.to.save = c('alpha',
                                       'beta',
                                       'sigma'),
                n.iter = 5000,
                n.thin = 5,
                model.file = textConnection(jags_code))
```
```{r, echo = FALSE, fig.height=5}
par(mfrow=c(1,3))
pars = jags_run$BUGSoutput$sims.list
for(i in c(1:2,4)) {
  hist(pars[[i]], breaks = 30, main = names(pars)[i], xlab = 'Parameter value')  
}
par(mfrow=c(1,1))
```

## Notes about OD Poisson model

- The way to think about OD models is via the data generating process. 

- We could compare this model to one without over dispersion via DIC (or if time, cross validation). We should also compute a posterior predictive distribution for full comparison

- In general, the parameter values (i.e. alpha and beta) tend to be more uncertain when you add in over dispersion

- Also in the data set is a variable called `dur` which represents how long they spent looking for the birds. This could be added in as an offset via the likelihood:
```
y[i] ~ dpois(dur[i] * exp(log_lambda[i]))
```

## Type 3: $t$-distributed HGLMs

- How do Bayesians deal with outliers?

- A common view is that we should delete these observations before we run the model, but what if we can't find a reason for doing so

- A good Bayesian will include outliers as part of the model. 

- One way of doing this is by switching from a normal distribution to a $t$-distribution

## Normal vs $t$

```{r, fig.height=5}
curve(dnorm, from = -5, to = 5)
curve(dt(x, df = 1), add = TRUE, col = 'red')
curve(dt(x, df = 4), add = TRUE, col = 'blue')
```

## Polluted data

- Suppose we had some data which looked like this:

```{r, echo = FALSE, fig.height= 5}
N = 100
set.seed(123)
x = runif(N)
y = rt(N, df = 3)*0.8 + 2 - 2*x
plot(x, y)
```

There are a few observations here which look a bit odd

## JAGS code for a $t$-model

```{r}
jags_code = '
model{
  # Likelihood
  for(i in 1:N) {
    y[i] ~ dt(alpha + beta * (x[i] - mean(x)), 
                sigma, df[i])
    df[i] ~ dcat(p)
  }
  alpha ~ dnorm(0, 1^-2)
  beta ~ dnorm(0, 1^-2)
  sigma ~ dt(0,1,1)T(0,)
}
'
```

## Fitting the model

```{r, message=FALSE, results = 'hide'}
jags_run = jags(data = list(N = N,
                            p = rep(1,10)/10,
                            y = y,
                            x = x),
                parameters.to.save = c('alpha',
                                       'beta',
                                       'df'),
                model.file = textConnection(jags_code))
```

## Output from the model

\tiny
```{r, fig.height=5}
dfs = jags_run$BUGSoutput$median$df
pars = jags_run$BUGSoutput$mean
cols = rainbow(10)
plot(x, y, col = cols[dfs])
lines(x, as.numeric(pars$alpha) + 
        as.numeric(pars$beta)*(x - mean(x)))
```

## Prior distributions on the degrees of freedom

- Here I've set a prior distribution on the degrees of freedom parameter to be a categorical distribution with probabilities 0.1 for df = 1, 2, ..., 10
- Smaller values of `df` mean that a data point is more likely to be an outlier
- The categorical distribution automatically looks up the right `df` value for each probability
- This model is impossible to fit in Stan, because it contains a discrete parameter

## Type 4: Ordinal data HGLMs

- Often we have a response variable which is ordinal, e.g. disagree, neutral, agree, etc
- There are lots of different (and complicated) ways to model such data
- Perhaps the easiest is to think of it as a hierarchical model with 'cut-points' on a latent linear regression

## An ordinal model example

- Suppose $y_i = \{\mbox{disagree, neutral, agree} \}$ and we make it dependent on a latent continuous variable $z_i$, so that :

$$y_i = \left\{ \begin{array}{ll} \mbox{agree} & \mbox{if } z_i> 0.5 \\
\mbox{neutral} & \mbox{if } -0.5 < z_i \le 0.5 \\
\mbox{disagree} & \mbox{if } z_i \le -0.5 \end{array} \right.$$

- We then give $z_i$ a prior distribution, e.g. $N(\beta_0 + \beta_1 x_i, \sigma^2)$


## Fitting ordinal models in JAGS

```{r}
jags_code = '
model{
  # Likelihood
  for(i in 1:N) {
    z[i] ~ dnorm(alpha + beta * (x[i] - mean(x)), 
                    sigma^-2)
    y[i] ~ dinterval(z[i], cuts)
  }
  alpha ~ dnorm(0, 100^-2)  
  beta ~ dnorm(0, 100^-2)  
  sigma ~ dt(0, 10, 1)T(0, )
}
'
```

## Simulating some example data

```{r}
N = 100
alpha = -1
beta = 0.2
sigma = 0.51
set.seed(123)
x = runif(N, 0, 10)
cuts = c(-0.5, 0.5)
z = rnorm(N, alpha + beta * (x - mean(x)), sigma)
y = findInterval(z, cuts)
```

## Simulated data - plot

```{r, fig.height=5}
plot(x, z, col = y + 1)
```


## Fitting in JAGS - needs initial values

```{r, message=FALSE, results = 'hide'}
jags_inits = function() {
  z = runif(N, -0.5, 0.5)
  z[y==0] = runif(sum(y==0), -1, -0.5)
  z[y==2] = runif(sum(y==2), 0.5, 1)
  return(list(z = z))
}
jags_run = jags(data = list(N = N,
                            y = y,
                            x = x,
                            cuts = cuts),
                inits = jags_inits,
                parameters.to.save = c('alpha',
                                       'beta',
                                       'sigma'),
                model.file = textConnection(jags_code))
```

## Output

\small
```{r}
print(jags_run)
```

## Summary

- We have now seen a number of different types of hierarchical GLM
- Many of the ideas of hierarchical linear models transfer over, but we can explore richer behaviour with hierarchical GLMs
- These have all used the normal, binomial or Poisson distribution at the top level, and have allowed for over-dispersion, robustness, and ordinal data, to name just three

